Precise spiking motifs in neurobiological and neuromorphic data

### 2022-11-08 manuscript.tex
Update on Overleaf.

More generally, the idea of using the synchrony of co-activation in a cell assembly is reminiscent of the hypothesis that was formalized by~\citet{hebb_organization_1949}. 
This is also expressed in the idea that different cortical areas could achieve binding by synchrony~\citep{fries_mechanism_2005}. The synchronicity will generate rhythms at different range of frequencies, with spikes arriving at peak susceptibility (top of a cycle) or down. Such a theory has surprisingly been validated in EEG recordings to explain for example the continuous wagon-wheel illusion, i.e. the perceived reversal of the rotational movements of the spikes in a rotating wheel~\citep{vanrullen_continuous_2006}. More generally, it can be shown that the phase of alpha oscillations (about $10$ Hz) is causally linked with modulations of cortical excitability and with visual perception~\citep{dugue_phase_2011}. 
%
\subsection{A further hypothesis: traveling waves}
To further investigate the role of precise timing, let us also focus on the role of differential timings in an assembly of neurons. As we have seen, a visual feature will induce the firing of different cells at different latencies~\citep{celebrini_dynamics_1993}. Using intracellular recordings, it was shown that the response to a focal visual activation would elicit a latency basin, that is, a graded onset of the neural response from the most activated to neighboring neurons~\citep{bringuier_horizontal_1999}. In particular, it was shown that the network of horizontal connections within a cortical area is typically not myelinated and that this latency basin would be determined by the propagation speed within the area. For generic visual scenes, these processes would generate a complex interplay between the dynamics of the sensory signal and the spatio-temporal determinants of these interactions and may underlay anticipatory mechanisms in the primary visual cortex~\citep{benvenuti_anticipatory_2020,le_bec_horizontal_2022}.
%

@@

More recently, advanced recording techniques have enabled the identification of travelling waves of neural activity in different areas of the cortex~\citep{muller_cortical_2018}. Authors review these findings, consider the mechanisms by which travelling waves are generated, and evaluate their possible roles in cortical function. In particular, spontaneous traveling waves naturally emerge from horizontal fiber time delays and travel through locally asynchronous-irregular states~\citep{davis_spontaneous_2021}. Studies of sensory-evoked neuronal responses often focus on mean spike rates, with fluctuations treated as internally-generated noise. However, fluctuations of spontaneous activity, often organized as traveling waves, shape stimulus-evoked responses and perceptual sensitivity. The mechanisms underlying these waves are unknown. Further, it is unclear whether waves are consistent with the low rate and weakly correlated ``asynchronous-irregular'' dynamics observed in cortical recordings. Here, the authors describe a large-scale computational model with topographically-organized connectivity and conduction delays relevant to biological scales. They find that spontaneous traveling waves are a general property of these networks. The traveling waves that occur in the model are sparse, with only a small fraction of neurons participating in any individual wave. Consequently, they do not induce measurable spike correlations and remain consistent with locally asynchronous irregular states. Further, by modulating local network state, they can shape responses to incoming inputs as observed \textit{in vivo}. Such waves also occur in motor areas and~\citet{linden_movement_2021} have presented ensemble-recordings of neurons in the lumbar spinal cord that indicate that, rather than alternating, the population is performing a low-dimensional ``rotation'' in neural space, in which the neural activity is cycling through all phases continuously during the rhythmic behavior.

>>>>

More generally, the idea of using the synchrony of co-activation in a cell assembly is reminiscent of the hypothesis that was formalized by~\citet{hebb_organization_1949}. 
This is also expressed in the idea that different cortical areas could achieve binding by synchrony~\citep{fries_mechanism_2005}. The synchronicity will generate rhythms at different range of frequencies, with spikes arriving at peak susceptibility (top of a cycle) or down. Such a theory has surprisingly been validated in EEG recordings to explain for example the continuous wagon-wheel illusion, i.e. the perceived reversal of the rotational movements of the spikes in a rotating wheel~\citep{vanrullen_continuous_2006}. More generally, it can be shown that the phase of alpha oscillations (about $10$ Hz) is causally linked with modulations of cortical excitability and with visual perception~\citep{dugue_phase_2011}. 
%
\subsection{A further hypothesis: traveling waves}
%
To further investigate the role of precise timing, let us also focus on the role of differential timings in an assembly of neurons. As we have seen, a visual feature will induce the firing of different cells at different latencies~\citep{celebrini_dynamics_1993}. Using intracellular recordings, it was shown that the response to a focal visual activation would elicit a latency basin, that is, a graded onset of the neural response from the most activated to neighboring neurons~\citep{bringuier_horizontal_1999}. In particular, it was shown that the network of the so-called horizontal connections within a cortical area is typically not myelinated and that this latency basin would be determined by the propagation speed within that area. For generic visual scenes, these processes would generate a complex interplay between the dynamics of the sensory signal and the spatio-temporal determinants of these interactions and may underlay anticipatory mechanisms in the primary visual cortex~\citep{benvenuti_anticipatory_2020,le_bec_horizontal_2022}.
%

@@

More recently, novel multi-unit recording techniques have enabled the identification of travelling waves of neural activity in different areas of the cortex~\citep{muller_cortical_2018}. Authors review these findings, consider the mechanisms by which travelling waves are generated, and evaluate their possible roles in cortical function. In particular, spontaneous traveling waves naturally emerge from horizontal fiber time delays and travel through locally asynchronous-irregular states~\citep{davis_spontaneous_2021}. Studies of sensory-evoked neuronal responses often focus on mean spike rates, with fluctuations treated as internally-generated noise. However, fluctuations of spontaneous activity, often organized as traveling waves, shape stimulus-evoked responses and perceptual sensitivity. The mechanisms underlying these waves are unknown. Further, it is unclear whether waves are consistent with the low rate and weakly correlated ``asynchronous-irregular'' dynamics observed in cortical recordings. Here, the authors describe a large-scale computational model with topographically-organized connectivity and conduction delays relevant to biological scales. They find that spontaneous traveling waves are a general property of these networks. The traveling waves that occur in the model are sparse, with only a small fraction of neurons participating in any individual wave. Consequently, they do not induce measurable spike correlations and remain consistent with locally asynchronous irregular states. Further, by modulating local network state, they can shape responses to incoming inputs as observed \textit{in vivo}. Such waves also occur in motor areas and~\citet{linden_movement_2021} have presented ensemble-recordings of neurons in the lumbar spinal cord that indicate that, rather than alternating, the population is performing a low-dimensional ``rotation'' in neural space, in which the neural activity is cycling through all phases continuously during the rhythmic behavior.
